---
title: "Wolfcamp data: IDW vs Multilevel B-Splines"
format:
  html:
    code-fold: false
    toc: true
    embed-resources: true
engine: knitr
execute:
  warning: false
  message: false
---

## Installation of the packages

The packages only need to be installed once, so this chunk is not evaluated when rendering.

```{r}
#| label: install
#| eval: false
install.packages("geoR")
install.packages("sf")
install.packages("gstat")
install.packages("MBA")
```

## Loading the data

```{r}
#| label: load-geoR
library(geoR)
data(wolfcamp)
```

We got wolfcamp data. In a first time, we will analyse the data in order to understand their repartition.

```{r}
#| label: summary-plot
summary(wolfcamp)

plot(wolfcamp)
```

Then, we create a csv file so as to stock the data.

```{r}
#| label: export-csv
wolfcamp_df <- data.frame(
  x = wolfcamp$coords[,1], 
  y = wolfcamp$coords[,2], 
  niveau_eau = wolfcamp$data
)

write.csv(wolfcamp_df, file = "wolfcamp_data.csv", row.names = FALSE)

head(wolfcamp_df)
```

Link to discover the first discussion : <https://share.gemini.google/mQVoQ6oey1LJ>

## Choice of the power parameter for IDW

Now, we choose the parameter p which allows to have the best predictions with the inverse distance weighted estimator.

```{r}
#| label: idw-optimization
# 1. Load required libraries
library(sf)
library(gstat)

# 1. Read the CSV file into an R data.frame
wolfcamp <- read.csv("wolfcamp_data.csv")

# 2. Convert the loaded data.frame into a spatial (sf) object

wolfcamp_sf <- st_as_sf(wolfcamp, coords = c("x", "y"))

# 3. Define a sequence of power parameters (p) to test
p_values <- seq(0.5, 10.0, by = 0.25)

# Initialize an empty vector to store the MSE for each p
mse_results <- numeric(length(p_values))

# 4. Outer Loop: Iterate over each possible parameter p
for (i in seq_along(p_values)) {
  
  current_p <- p_values[i]
  squared_errors <- numeric(nrow(wolfcamp_sf))
  
  # 5. Inner Loop: Leave-One-Out Cross-Validation
  for (j in 1:nrow(wolfcamp_sf)) {
    
    # STRICT SEPARATION: Remove test point 'j' from training data
    train_data <- wolfcamp_sf[-j, ]
    test_data  <- wolfcamp_sf[j, ]
    
    # Fit IDW model on training data and predict on the test point
    # 'idp' is the gstat argument for the power parameter
    idw_fit <- idw(
      formula = niveau_eau ~ 1,
      locations = train_data,
      newdata = test_data,
      idp = current_p,
      debug.level = 0 # Suppress console output for speed
    )
    
    # Extract prediction and calculate squared error
    predicted_val <- idw_fit$var1.pred
    actual_val <- test_data$niveau_eau
    
    squared_errors[j] <- (predicted_val - actual_val)^2
  }
  
  # 6. Calculate the Mean Squared Error for the current p
  mse_results[i] <- mean(squared_errors)
}

# 7. Identify the optimal p
results_df <- data.frame(Power = p_values, MSE = mse_results)
optimal_p <- results_df$Power[which.min(results_df$MSE)]

print(results_df)
cat("\nThe optimal power parameter (p) is:", optimal_p, "\n")
# Plot the evolution of MSE vs. p
plot(results_df$Power, results_df$MSE, type = "b", pch = 19, col = "blue",
     xlab = "Power Parameter (p)", ylab = "Mean Squared Error (MSE)",
     main = "IDW Optimization: MSE vs. Power Parameter")

# Add a vertical red dashed line to highlight the optimal p
abline(v = optimal_p, col = "red", lty = 2, lwd = 2)
```

## Comparison of IDW and MBA

```{r}
#| label: idw-vs-mba
library(sf)
library(gstat)
library(MBA)

# ---------------------------------------------------------
# PART 1: Define the 20x20 Grid
# ---------------------------------------------------------
# Create a sequence of 20 evenly spaced X and Y coordinates
x_seq <- seq(min(wolfcamp$x), max(wolfcamp$x), length.out = 20)
y_seq <- seq(min(wolfcamp$y), max(wolfcamp$y), length.out = 20)

# Combine them into a grid of 400 points
pred_grid <- expand.grid(x = x_seq, y = y_seq)
grid_sf <- st_as_sf(pred_grid, coords = c("x", "y"))

# ---------------------------------------------------------
# PART 2: Interpolate on the Grid
# ---------------------------------------------------------
# A. IDW Interpolation on the grid (using your optimal_p from earlier)
idw_grid_result <- idw(
  formula = niveau_eau ~ 1, 
  locations = wolfcamp_sf, 
  newdata = grid_sf, 
  idp = 3.75
)

# B. MBA Interpolation on the grid
# mba.surf takes an [x, y, z] matrix and grid dimensions
xyz_data <- wolfcamp[, c("x", "y", "niveau_eau")]
mba_grid_result <- mba.surf(xyz_data, no.X = 20, no.Y = 20, extend = TRUE)

# ---------------------------------------------------------
# PART 3: Calculate a "Proper Score" (LOOCV RMSE)
# ---------------------------------------------------------

# A. Score IDW (using gstat's built-in krige.cv)
idw_cv <- krige.cv(niveau_eau ~ 1, wolfcamp_sf, set = list(idp = optimal_p), debug.level = 0)
idw_rmse <- sqrt(mean(idw_cv$residual^2))

# B. Score MBA (Manual LOOCV using mba.points)
mba_residuals <- numeric(nrow(wolfcamp))

for (i in 1:nrow(wolfcamp)) {
  # Separate train and test
  train_xyz <- xyz_data[-i, ]
  test_xy <- xyz_data[i, c("x", "y")]
  
  # Predict the left-out point using mba.points
  mba_pred <- mba.points(train_xyz, test_xy)$xyz.est
  
  # Calculate residual (Actual - Predicted)
  actual_val <- xyz_data$niveau_eau[i]
  predicted_val <- mba_pred[1, "z"]
  
  mba_residuals[i] <- actual_val - predicted_val
}

mba_rmse <- sqrt(mean(mba_residuals^2))

# ---------------------------------------------------------
# PART 4: Compare
# ---------------------------------------------------------
cat("Proper Score Comparison (Root Mean Squared Error):\n")
cat("IDW RMSE: ", idw_rmse, "\n")
cat("MBA RMSE: ", mba_rmse, "\n\n")

if(idw_rmse < mba_rmse) {
  cat("Conclusion: IDW is the more accurate method for this dataset.\n")
} else {
  cat("Conclusion: Multilevel B-Splines (MBA) is the more accurate method for this dataset.\n")
}
```
